---
title: "Vowel formants example"
author: "Steven Moran"
date: "(`r format(Sys.time(), '%d %B, %Y')`)"
output:
  github_document:
      toc: true
---

# Getting started

## Load the R packages that we need for analysis

In R, you need to install (once) any R software package (aka library) that you want to use, before you load it with the `library()` function.

This code will check whether or not you have those packages, and if not, will install them and load the packages for you.

```{r, message=FALSE}
if(!require("tidyverse")) install.packages("tidyverse")
if(!require("ggpubr")) install.packages("ggpubr")
if(!require("devtools")) install.packages("devtools")
library(devtools)
if(!require("ggConvexHull")) devtools::install_github("cmartin/ggConvexHull")
```

Once a package is already installed on your computer, you can simply load it with the `library()` function, like this:

```{r, message=FALSE}
library(tidyverse)
library(ggpubr)
library(ggConvexHull)
```

If you want to know what these do, see the notes section at the end of this file.


## Load some example vowel data

First make sure that you set your working directory to where this file is by clicking on RStudio > Session > Set Working Directory > To Source File Location.

Then we can load some example data in the same directory as this file.

```{r, message=FALSE}
df <- read_csv('data.csv')
```

Note that this is an example, and what you really want to be able to do is load the data that we collect in class from in the Google spreadsheet.

To do so, first download it from the shared Google sheets as a [CSV file](https://en.wikipedia.org/wiki/Comma-separated_values). 

Put it in the same directory as this file.

Then you can load it with the same command above.

## Have a look at the raw data

Let's have a look at the data. 

```{r}
df
```

# Working with the data

## Overview

We had a quick look at the raw data. You can also do directly in RStudio by clicking on the dataframe in the Environment tab.

There are lots of ways of doing preliminary data analysis and one great way is to visualize the data!

Recall our discussion of how the International Phonetic Alphabet (IPA) [vowel chart](https://en.wikipedia.org/wiki/Vowel_diagram) looks -- head faces left, vertical access is the jaw's height (closed to open) and the horizontal access is the position of the tongue (front to back in the mouth).


## Single speaker

Let's look at one speaker in our sample. This is where the data manipulation R package [dplyr](https://dplyr.tidyverse.org) comes in handy!

The dplyr package has several functions that allow us to `filter()` rows and to `select()` columns (among many other useful things!).

Let's filter out a single speaker and plot their vowels. You can set the ID to yourself!

```{r}
single_speaker <- df %>% filter(ID == 1)
```

Above, we have saved the output of filtering by `ID == 1` (read "subject ID equals one") to a new data frame and named it `single_speaker`. Let's have a look.

```{r}
single_speaker
```

Now let's plot Subject ID 1's vowel formant values for F1 and F2.

```{r}
ggplot(single_speaker, aes(x = F2, y = F1, color = Vowel)) + 
  geom_convexhull(alpha = 0, colour = "black") +
  geom_text(aes(label = Vowel), size = 12) +
  scale_x_reverse() +
  scale_y_reverse() +
  coord_cartesian() +
  theme_pubr(border = TRUE, legend = "none") +
  theme(axis.ticks.length = unit(.25, "cm")) +
  scale_color_manual(
    name = "Vowel",
    values = c(
      "a" = "indianred",
      "i" = "black",
      "u" = "cornflowerblue",
      "o" = "wheat3",
      "e" = "purple3"
    )
  ) +
  ggtitle('Subject ID 1')
```

We can also get a summary of the data.

```{r}
summary(df)
```

## All speakers

Let's plot all of the F1 and F2 formants in our sample and see what they look like. We use the `ggplot()` function with lots of other functions. This is complicated -- do not worry, we will go through this mess in class.

```{r}
ggplot(df, aes(x = F2, y = F1, color = Vowel)) +
  geom_convexhull(alpha = 0, colour = "purple") +
  scale_x_reverse() +
  scale_y_reverse() +
    coord_cartesian() +
  theme_pubr(border = TRUE, legend = "none") +
  theme(axis.ticks.length = unit(.25, "cm")) +
  scale_color_manual(
    name = "Vowel",
    values = c(
      "a" = "indianred",
      "i" = "black",
      "u" = "cornflowerblue",
      "o" = "wheat3",
      "e" = "purple3"
    )
  ) +
  ggtitle('All speakers')


```


```{r}
ggplot(df, aes(x = F2, y = F1, color = Vowel)) + 
  geom_convexhull(alpha = 0, colour = "black") +
  geom_text(aes(label = Vowel), size = 12) +
  scale_x_reverse() +
  scale_y_reverse() +
  coord_cartesian() +
  theme_pubr(border = TRUE, legend = "none") +
  theme(axis.ticks.length = unit(.25, "cm")) +
  scale_color_manual(
    name = "Vowel",
    values = c(
      "a" = "indianred",
      "i" = "black",
      "u" = "cornflowerblue",
      "o" = "wheat3",
      "e" = "purple3"
    )
  ) +
  ggtitle('All speakers')
```

For each speaker, we can add the vowel polygon lines.

```{r}
ggplot(df, aes(x = F2, y = F1)) +
  geom_convexhull(alpha = 0, aes(colour = ID)) +
  geom_text(aes(label = Vowel), size = 10) +
  scale_x_reverse() +
  scale_y_reverse() +
  coord_cartesian() +
  theme_pubr(border = TRUE) +
  theme(axis.ticks.length = unit(.25, "cm"))
```


# Analyses

## Phonetic differences by sex

Let's plot individual vowel polygons by the reported sex of the speaker. As we expect from the literature on [phonetic differences between male and female speech](https://compass.onlinelibrary.wiley.com/doi/abs/10.1111/j.1749-818X.2009.00125.x), the vowel polygon of the women in our sample displays greater acoustic range than the men.

```{r}
ggplot(df, aes(x = F2, y = F1)) +
  geom_convexhull(alpha = 0, aes(colour = Sex)) +
  geom_text(aes(label = Vowel), size = 10) +
  scale_x_reverse() +
  scale_y_reverse() +
  coord_cartesian() +
  theme_pubr(border = TRUE) +
  theme(axis.ticks.length = unit(.25, "cm")) +
  theme_bw()
```

We can also plot individual vowel polygon spaces by sex and filled in by color. 

```{r}
ggplot(df, aes(x = F2, y = F1)) +
  geom_convexhull(alpha = 0.3, lwd = 0, aes(colour = ID, fill = Sex)) +
  geom_text(aes(label = Vowel), size = 10) +
  scale_x_reverse() +
  scale_y_reverse() +
  coord_cartesian() +
  theme_pubr(border = TRUE) +
  theme(axis.ticks.length = unit(.25, "cm"))
```

## Differences in pitch

Recall that we also recorded our F0 -- our fundamental frequency, which we perceive as our pitch. This is in the F0 column of our data frame:

```{r}
head(df)
```

Again, we access any column in the data frame with the `$` operator with the name of the column as the suffix, e.g. let's get all values (aka observations, rows) of F0:

```{r}
df$F0
```

We can also use various R functions on columns. For example, you can summarize numerical values from a column with the `summary()` function!

```{r}
summary(df$F0)
```

Perhaps we want to summarize the F0 by sex? In "[base R](https://iqss.github.io/dss-workshops/R/Rintro/base-r-cheat-sheet.pdf)":

```{r}
# Get a column by name
df$F0

# Get rows by filtering the contents of a colum; here by "sex is F(emale)"
df[df$Sex == "F", ]

# Get just the contents of the column we want by the filter -- ugh this is painful!
df[df$Sex == "F", ]$F0

# Now let's summarize that mess
summary(df[df$Sex == "F", ]$F0)

# And the men
summary(df[df$Sex == "M", ]$F0)
```

## Pitch versus height

Lastly, let's see if there is a correlation between one's height and one's F0.

First, let's get the data we want by "grouping" the height variable and summarizing the mean F0 for each by height.

```{r}
tmp <- df %>% group_by(Height) %>% summarize(Mean_F0 = mean(F0))
```

Now let's plot the results as a linear regression.

```{r}
ggplot(tmp, aes(x=Height, y=Mean_F0)) +
  geom_point()
```


```{r}
ggplot(tmp, aes(Height, Mean_F0)) +
  geom_point(colour = "black", alpha = 1) +
  xlab("Height") +
  ylab("Mean F0") +
  theme_pubr(border = TRUE, margin = TRUE) + 
  geom_smooth(method = "lm", se = FALSE, colour = "red") +
  ggpubr::stat_cor(aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~")))
```

We can also get the model statistics directly from the data with these functions.

```{r}
model1 <- lm(Mean_F0 ~ Height, data = tmp)
summary(model1)
```


# Notes

Some notes on what these libraries do (you can skip this and go to the next section):

The [tidyverse](https://www.tidyverse.org) package will install several [R packages](https://www.tidyverse.org/packages/) including:

* [ggplot](https://ggplot2.tidyverse.org) for creating nice looking plots
* [dplyr](https://dplyr.tidyverse.org) for manipulating data easily

We are going to additionally use [ggpubr](http://www.sthda.com/english/wiki/ggpubr-create-easily-publication-ready-plots) package because it is helpful in making [professional looking plots](https://medium.com/swlh/beautiful-charts-with-r-and-ggpubr-c94122d6b7c6). 

And the [ggConvexHull](https://github.com/cmartin/ggConvexHull) package, which extends the `geom_polyon()` function in the ggplot package, so that we can add a [convex hull](https://en.wikipedia.org/wiki/Convex_hull) around our vowel polygon data. This package is available via GitHub and hence the [devtools](https://devtools.r-lib.org) package.
